Weighing Chances and Use Against Risks and Misuse of Emerging Genome Analysis and Manipulation Technology

### 2022-05-31 draft.tex
minimal changes to citation

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artifically pollinated date trees\cite[p.633]{mayr}.
Since then several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic matieral to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a seperate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.
The most direct and controlled form of genetic modifiction is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

@@

\bibitem{abelsidentity} Heiz Abels,
\textbf{Identität} (July 15th, 2010)
https://books.google.de/books?hl=de\&lr=\&id=Z0z4y6WUZLgC\&oi=fnd\&pg=PA13\&dq=identit\%C3\%A4t\&ots=wmnh96n2l2\&sig=FME9rlt1rE9DFFGSZz60R\_F5e-M\&redir\_esc=y\#v=onepage\&q=identit\%C3\%A4t\&f=false

@@

\bibitem{apgfk} GfK Public Affairs \& Corporate Communications
\textbf{THE AP-GfK POLL, A survey of the American general population (ages 18+)} (March, 2014)
http://surveys.associatedpress.com/data/GfK/AP-GfK\%20March\%202014\%20Poll\%20Topline\%20\%20Final\_SCIENCE.pdf

@@

\bibitem{companydata} Laestadius, L., Rich, J. \& Auer, P
\textbf{All your data (effectively) belong to us: data practices among direct-to-consumer genetic testing firms} (September 22th, 2016)
https://www.nature.com/articles/gim2016136\#citeas

@@

\bibitem{protectionoffreedomsact} Aaron Opoku Amankwaa, Carole McCartney,
\textbf{The UK National DNA Database: Implementation of the Protection of Freedoms Act 2012} (March, 2018)
https://www.sciencedirect.com/science/article/abs/pii/S0379073817305571?via \%3Dihub

@@

\bibitem{nordgrenjuengst} Anders Nordgren, E.T. Juengst,
\textbf{Can genomics tell me who I am? Essentialistic rhetoric in direct-to-consumer DNA testing} (May 28th, 2009)
https://www.tandfonline.com/doi/citedby/10.1080/14636770902901595?scroll=top\&needAccess=true

>>>>

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artifically pollinated date trees\cite{mayr}.
Since then several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic matieral to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a seperate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.
The most direct and controlled form of genetic modifiction is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

\subsection{Genetic Privacy}

@@

\bibitem{abelsidentity} Heiz Abels,
\textbf{Identität} (July 15th, 2010)
https://link.springer.com/content/pdf/10.1007/978-3-531-90437-5.pdf

@@

\bibitem{apgfk} GfK Public Affairs \& Corporate Communications
\textbf{THE AP-GfK POLL, A survey of the American general population (ages 18+)} (March, 2014)
http://surveys.associatedpress.com/data/
GfK/AP-GfK\%20March\%202014
\%20Poll\%20Topline\%20\%
20Final\_SCIENCE.pdf

@@

\bibitem{companydata} Laestadius, L., Rich, J. \& Auer, P
\textbf{All your data (effectively) belong to us: data practices among direct-to-consumer genetic testing firms} (September 22th, 2016)
https://www.nature.com/articles/gim2016136

@@

\bibitem{protectionoffreedomsact} Aaron Opoku Amankwaa, Carole McCartney,
\textbf{The UK National DNA Database: Implementation of the Protection of Freedoms Act 2012} (March, 2018)
https://www.sciencedirect.com/science/article/abs/pii/
S0379073817305571?via \%3Dihub

@@

\bibitem{nordgrenjuengst} Anders Nordgren, E.T. Juengst,
\textbf{Can genomics tell me who I am? Essentialistic rhetoric in direct-to-consumer DNA testing} (May 28th, 2009)
https://www.tandfonline.com/doi/citedby/10.1080/
14636770902901595?scroll=top\&needAccess=true
